---
title: "Predict race/ethnicity based on census data"
output: html_document
---

## Setups

```{r}
if (!requireNamespace('wru')) install.packages('wru')
# rm(list = ls())

library(wru)
library(tidyverse)
library(lubridate)
source('utils/r-utils.R')
```

## Load data
and use the `wru` package to predict race based on last_names only

```{r}
pubmed_race_df <- read_tsv('data/pubmed/authors.tsv.xz') %>%
  filter(reverse_position == 1) %>%
  left_join(read_tsv('data/pubmed/articles.tsv.xz'), by = 'pmid') %>%
  mutate(year = substr(publication_date, 1, 4) %>% ymd(truncated = 2)) %>%
  rename('surname' = last_name) %>% 
  predict_race(surname.only = T, impute.missing = F) %>%
  mutate(publication_date = ymd(publication_date, truncated = 2))

ismb_race_df <- read_tsv('data/ismb/keynotes.tsv') %>%
  rename('surname' = last_name) %>%
  mutate(year = ymd(year, truncated = 2)) %>% 
  predict_race(surname.only = T, impute.missing = F)

alpha <- qnorm(0.975)
n_years <- 18
my_jours <- unique(pubmed_race_df$journal)
n_jours <- length(my_jours)
```

Pubmed: 7587 last_names out of 28436 were not matched.
ISMB: 29 surnames out of 125 were not matched.

## Descriptive statistics
Prepare data frames for later analyses:

- rbind results of race predictions in ISMB and Pubmed
- pivot long
- compute mean, sd, marginal error
- add "White" vs. "Others"

```{r}
ismb_pubmed <- ismb_race_df %>%
  select(year, contains('pred')) %>%
  mutate(publication_date = year,
         type = 'ISMB',
         journal = 'ISMB') %>%
  rbind(
    pubmed_race_df %>%
      select(year, journal, publication_date, contains('pred')) %>%
      mutate(type = 'Pubmed')
  ) %>%
  pivot_longer(contains('pred'),
               names_to = 'Race',
               values_to = 'Probabilities') %>%
  recode_race() %>%
  group_by(year, Race, type, publication_date) %>%
  add_count() %>%
  mutate(
    mean_prob = mean(Probabilities, na.rm = T),
    sd_prob = sd(Probabilities, na.rm = T),
    n = mean(n),
    me_prob = alpha * sd_prob / sqrt(n)
  ) %>%
  ungroup() %>%
  mutate(combined_race = ifelse(Race == 'White', 'White', 'Others'))

```

Stacked bar plots:
```{r}
plot_dynamic_race <- vector('list', length = n_jours)
i <- 0
for (jour in my_jours){
  i <- i + 1
  plot_dynamic_race[[i]] <- ismb_pubmed %>%
  filter(type == 'Pubmed') %>%
    filter(journal == jour) %>%
    group_by(year, Race) %>%
    summarise(mean_prob = mean(Probabilities, na.rm = T)) %>%
    dynamic_contribution(jour)
}

do.call(cowplot::plot_grid, c(plot_dynamic_race, nrow = 3))

ismb_pubmed %>%
  filter(type == 'ISMB') %>% 
  group_by(year, Race) %>%
  summarise(mean_prob = mean(Probabilities, na.rm = T)) %>%
  dynamic_contribution('ISMB keynote speakers')
```


```{r}
# boxplot ismb, pubmed faceted by year
ismb_pubmed %>% 
  filter(Race == 'White') %>% 
  ggplot(aes(x = type, fill = type, y = Probabilities)) +
  geom_violin(alpha = 0.8) +
  facet_wrap(~ year(year), nrow = 3) +
  theme_bw() +
  labs(x = NULL, y = 'Probabilities of being white') +
  scale_x_discrete(label = NULL) +
  theme(legend.title = element_blank()) +
  scale_fill_viridis_d()
```

Confidence interval of the mean? (Aggregation)

Null: The proportion of white keynote ISMB speakers is not larger the proportions of white last authors published in Pubmed Central.
Alternative: The proportion of white keynote ISMB speakers is larger the proportions of white last authors published in Pubmed Central.
How do we take time component into account?
Right now, I'm thinking we can use `year` as a random variable.
Perhaps we should allow for random slope as well?

Bootstrap?

```{r}
get_summary <- function(df){
  df %>% filter(type == 'ISMB') %>% select(- Probabilities) %>% distinct()
}

ggplot(ismb_pubmed %>% filter(Race == 'White' | Race == 'Asian'),
       aes(x = year)) +
  facet_grid(cols = vars(Race)) +
  geom_smooth(aes(x = publication_date, y = Probabilities, color = type)) +
  geom_point(
    data = . %>% filter(type == 'ISMB'),
    aes(y = Probabilities, color = type),
    alpha = 0.2,
    show.legend = F
  ) +
  geom_point(
    data = . %>% get_summary(),
    aes(y = mean_prob, color = type),
    shape = 2,
    alpha = 0.5,
    show.legend = F
  ) +
  geom_linerange(
    data = . %>% get_summary(),
    aes(
      ymin = mean_prob - me_prob,
      ymax = mean_prob + me_prob,
      color = type
    ),
    alpha = 0.5,
    show.legend = F
  ) +
  theme_bw() +
  theme(legend.title = element_blank(),
        legend.position = c(0.1, 0.85)) +
  scale_color_viridis_d(option = 'E') +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = NULL)

```


