---
title: "Predict race/ethnicity based on census data"
output: html_document
---

## Setups

```{r}
if (!requireNamespace('wru')) install.packages('wru')
rm(list = ls())

library(wru)
library(tidyverse)
library(lubridate)
source('utils/r-utils.R')
```

## Load data
and use the `wru` package to predict race based on last_names only

```{r}
pubmed_race_df <- read_tsv('data/pubmed/authors.tsv.xz') %>%
  filter(reverse_position == 1) %>%
  left_join(read_tsv('data/pubmed/articles.tsv.xz'), by = 'pmid') %>%
  mutate(year = substr(publication_date, 1, 4) %>% ymd(truncated = 2)) %>%
  rename('surname' = last_name) %>% 
  predict_race(surname.only = T, impute.missing = F) %>%
  mutate(publication_date = ymd(publication_date, truncated = 2))

ismb_race_df <- read_tsv('data/ismb/keynotes.tsv') %>%
  rename('surname' = last_name) %>%
  mutate(year = ymd(year, truncated = 2)) %>% 
  predict_race(surname.only = T, impute.missing = F)

n_years <- 18
my_jours <- unique(pubmed_race_df$journal)
n_jours <- length(my_jours)
```

Pubmed: 7587 last_names out of 28436 were not matched.
ISMB: 29 surnames out of 125 were not matched.

## Descriptive statistics

```{r}
pubmed_race_plot <- pubmed_race_df %>%
  dplyr::select(year, journal, contains('pred')) %>%
  pivot_longer(
    - c(journal, year), 
    names_to = 'Race', 
    values_to = 'Probabilities') %>%
  recode_race()

ismb_race_plot <- ismb_race_df %>%
  dplyr::select(year, contains('pred')) %>%
  pivot_longer(
    - c(year),
    names_to = 'Race',
    values_to = 'Probabilities') %>%
  recode_race()

# Check: sum of all average probabilities for each year should add up to 1
pubmed_race_plot %>%
  group_by(year, Race) %>%
  summarise(mean_prob = mean(Probabilities, na.rm = T)) %>%
  summarise(all_prob = sum(mean_prob))
```

Stacked bar plots:
```{r}
plot_dynamic_race <- vector('list', length = n_jours)
i <- 0
for (jour in my_jours){
  i <- i + 1
  plot_dynamic_race[[i]] <- pubmed_race_plot %>%
    filter(journal == jour) %>%
    group_by(year, Race) %>%
    summarise(mean_prob = mean(Probabilities, na.rm = T)) %>%
    dynamic_contribution(jour)
}

do.call(cowplot::plot_grid, c(plot_dynamic_race, nrow = 3))

ismb_race_plot %>%
  group_by(year, Race) %>%
  summarise(mean_prob = mean(Probabilities, na.rm = T)) %>%
  dynamic_contribution('ISMB keynote speakers')
```


```{r}
alpha <- qnorm(0.975)
ismb_pubmed <- pubmed_race_plot %>% 
  select(- journal) %>% 
  mutate(type = 'Pubmed') %>% 
  rbind(ismb_race_plot %>% mutate(type = 'ISMB'))

# boxplot ismb, pubmed faceted by year
ismb_pubmed %>% 
  filter(Race == 'White') %>% 
  ggplot(aes(x = type, fill = type, y = Probabilities)) +
  geom_violin(alpha = 0.8) +
  facet_wrap(~ year(year), nrow = 3) +
  theme_bw() +
  labs(x = NULL) +
  scale_x_discrete(label = NULL) +
  theme(legend.title = element_blank()) +
  scale_fill_viridis_d()
```

Confidence interval of the mean? (Aggregation)

Null: The proportion of white keynote ISMB speakers is not larger the proportions of white last authors published in Pubmed Central.
Alternative: The proportion of white keynote ISMB speakers is larger the proportions of white last authors published in Pubmed Central.
How do we take time component into account?
Right now, I'm thinking we can use `year` as a random variable.
Perhaps we should allow for random slope as well?

Bootstrap?

```{r}
pub_ci <- pubmed_race_df %>%
  dplyr::select(year, journal, publication_date, contains('pred')) %>%
  pivot_longer(-c(journal, year, publication_date),
               names_to = 'Race',
               values_to = 'Probabilities') %>%
  recode_race() %>%
  filter(Race == my_race)

my_race <- 'White'

irace_mean_sd <- ismb_pubmed %>% 
  filter(Race == my_race) %>%
  group_by(year, Race, type) %>%
  add_count() %>%
  summarise(mean_prob = mean(Probabilities, na.rm = T),
            sd_prob = sd(Probabilities, na.rm = T),
            n = mean(n),
            me_prob = alpha*sd_prob/sqrt(n))

ggplot() +
  geom_smooth(data = pub_ci,
              aes(x = publication_date, y = Probabilities),
              color = '#eae428ff',
              fill = 'grey85') +
  geom_point(data = filter(ismb_race_plot, Race == my_race),
             aes(x = year, y = Probabilities),
             alpha = 0.2, color = '#1f968bff') +
  geom_smooth(data = filter(ismb_race_plot, Race == my_race),
             aes(x = year, y = Probabilities),
             alpha = 0.2, color = '#1f968bff') +
  geom_point(data = filter(irace_mean_sd, type == 'ISMB'), 
             aes(x = year, y = mean_prob),
             shape = 2,
             alpha = 0.5) +
  geom_linerange(data = filter(irace_mean_sd, type == 'ISMB'), 
                aes(x = year,
                    ymin = mean_prob - me_prob,
                    ymax =mean_prob + me_prob),
                alpha = 0.5) +
  theme_bw() +
  scale_color_viridis_d(option = 'E') +
  labs(x = NULL)

```

