---
title: "Visualize gender predictions based on census data"
output: html_document
---

## Setups

```{r message = F}
library(tidyverse)
library(lubridate)
source('utils/r-utils.R')
```

## Load data

```{r}
alpha <- qnorm(0.975)

simple_fore_names <- read_tsv('data/names/simple-fore-names.tsv.xz')
gender_df <- read_tsv('data/gender/genderize.tsv')

pubmed_df <- read_tsv('data/pubmed/authors.tsv.xz') %>%
  filter(reverse_position == 1) %>%
  left_join(read_tsv('data/pubmed/articles.tsv.xz'), by = 'pmid') %>%
  mutate(year = substr(publication_date, 1, 4) %>% ymd(truncated = 2),
         publication_date = publication_date %>% ymd(truncated = 2)) %>% 
  left_join(simple_fore_names %>% select(- n_authors), by = 'fore_name') %>% 
  left_join(gender_df, by = 'fore_name_simple') %>% 
  mutate(final_gender_pred = ifelse(probability_male > 0.5, 'Male', 'Female'))

iscb_df <- read_tsv('data/iscb/keynotes.tsv') %>%
  mutate(publication_date = ymd(year, truncated = 2),
         year = ymd(year, truncated = 2)) %>% 
  left_join(simple_fore_names %>% select(- n_authors), by = 'fore_name') %>% 
  left_join(gender_df, by = 'fore_name_simple') %>% 
  mutate(final_gender_pred = ifelse(probability_male > 0.5, 'Male', 'Female'))

start_year <- 1997
end_year <- 2020
n_years <- end_year - start_year
my_jours <- unique(pubmed_df$journal)
my_confs <- unique(iscb_df$Conference)
n_jours <- length(my_jours)
n_confs <- length(my_confs)
```

Number of speakers/fellows in compbio conferences across years:

```{r}
iscb_df %>%
  select(year, Conference) %>% 
  count(year, Conference) %>% 
  ggplot(aes(x = year(year), y = n, color = Conference)) +
  geom_point() +
  theme_bw() +
  geom_line(alpha = 0.5) +
  coord_cartesian(ylim = c(0, 13)) +
  scale_x_continuous(breaks = seq(1998, 2020, 2)) +
  scale_y_continuous(breaks = seq(0, 14, 2)) +
  scale_color_viridis_d() +
  labs(x = 'Year', y = 'Number of keynote speakers/fellows')
```

## Missing data

```{r}
iscb_df %>% filter(is.na(probability_male))
pubmed_df %>% filter(is.na(probability_male)) %>% sample_n(10)
sum(is.na(pubmed_df$probability_male))
```

2 iscb forenames not genderized.
2835 pubmed forenames not genderized.
These mostly include initials only, hyphenated names and perhaps names with accent marks.

```{r}
pubmed_df %>% filter(!is.na(probability_male)) %>% count(journal, year)
pubmed_df %>% filter(is.na(probability_male)) %>% 
  count(year, journal) %>% 
  ggplot(aes(year(year), n, fill = journal)) +
  geom_bar(stat = 'identity') +
  scale_fill_viridis_d() +
  ylab('Number of forenames NOT genderized')
```

The metadata for Bioinformatics papers before 2002 only have initials of first and/or middle names. 
Therefore, we do not have prediction for genders before this year.

```{r}
table(iscb_df$final_gender_pred) %>% 
  rbind(table(pubmed_df$final_gender_pred)) %>% 
  chisq.test()
```

## Descriptive statistics
Prepare data frames for later analyses:

- rbind results of race predictions in iscb and Pubmed
<!-- - pivot long -->
- compute mean, sd, marginal error

```{r}
iscb_pubmed <- iscb_df %>%
  rename('journal' = Conference) %>% 
  select(year, journal, probability_male, publication_date, final_gender_pred) %>%
  mutate(type = 'Keynote speakers/Fellows') %>%
  bind_rows(
    pubmed_df %>%
      select(year, journal, probability_male, publication_date, final_gender_pred) %>%
      mutate(type = 'Pubmed')
  ) %>%
  mutate(probability_female = 1 - probability_male) %>% 
  pivot_longer(contains('probability'),
               names_to = 'gender',
               values_to = 'probabilities') %>%
  group_by(type, year, publication_date, gender) %>%
  add_count() %>%
  mutate(
    mean_prob = mean(probabilities, na.rm = T),
    sd_prob = sd(probabilities, na.rm = T),
    n = mean(n),
    me_prob = alpha * sd_prob / sqrt(n)
  ) %>%
  ungroup() 
```

## Stacked bar plots:
### By journals

```{r}
pubmed_gender <- vector('list', length = n_jours)
i <- 0

for (jour in my_jours){
  i <- i + 1
  pubmed_gender[[i]] <- iscb_pubmed %>%
    filter(type == 'Pubmed' & journal == jour) %>%
    group_by(year, gender, journal) %>%
    summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
    ungroup() 
}

bind_rows(pubmed_gender) %>%
  gender_breakdown(start_year = 2002, end_year - 1, 'journal')

```

### By conference keynotes/fellows:
```{r fig.height=6}
i <- 0
iscb_gender <- vector('list', length = n_confs)

for (conf in my_confs){
  i <- i + 1
  iscb_gender[[i]] <- iscb_pubmed %>%
    filter(type != 'Pubmed' & journal == conf) %>%
    group_by(year, gender, journal) %>%
    summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
    ungroup()
}

bind_rows(iscb_gender) %>%
  gender_breakdown(start_year, end_year, 'journal')
```

### All fellows and keynotes compared to three ISCB partner journals:
```{r fig.width=10, fig.height=4}
iscb_pubmed %>%
  group_by(year, type, gender) %>%
  summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
  ungroup() %>%
  gender_breakdown(start_year, end_year, 'type', facet_by = 'col')
```

## Overview of gender prediction distribution

```{r}
iscb_pubmed %>% 
  filter(gender == 'probability_male') %>% 
  ggplot(aes(x = probabilities, y = ..density.., fill = type)) +
  geom_histogram(position = 'dodge') +
  scale_fill_viridis_d(option = 'D', end = 0.8) +
  guides(label = F) +
  theme(legend.title = element_blank()) +
  labs(x = 'probability_male')
```

## Mean and standard deviation of predicted probabilities

```{r fig.width=10, fig.height=4}
get_summary <- function(df){
  df %>% filter(type != 'Pubmed') %>% select(- probabilities) %>% distinct()
}

ggplot(iscb_pubmed %>% filter(gender == 'probability_male'),
       aes(x = year)) +
  # facet_grid(cols = vars(fct_rev(gender))) +
  geom_point(
    data = . %>% filter(type != 'Pubmed'),
    aes(y = probabilities, color = type),
    alpha = 0.2,
    show.legend = F
  ) +
  geom_point(
    data = . %>% get_summary(),
    aes(y = mean_prob, color = type),
    shape = 2,
    alpha = 0.5,
    show.legend = F
  ) +
  geom_linerange(
    data = . %>% get_summary(),
    aes(
      ymin = mean_prob - me_prob,
      ymax = mean_prob + me_prob,
      color = type
    ),
    alpha = 0.5,
    show.legend = F
  ) +
  geom_smooth(aes(x = publication_date, y = probabilities, color = type, fill = type), alpha = 0.2) +
  theme_bw() +
  scale_y_continuous(breaks = seq(0, 1, 0.2)) +
  theme(legend.title = element_blank()) +
  scale_color_viridis_d(option = 'D', end = 0.8) +
  scale_fill_viridis_d(option = 'D', end = 0.8) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = NULL, y = 'probability_male')
```


