---
title: "Summary statistics and missing data analysis"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_download: true
    highlight: tango
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(rnaturalearth)
library(wru)
source('utils/r-utils.R')
theme_set(theme_bw() + theme(legend.title = element_blank()))
```

## General data read-in

```{r}
load('Rdata/raws.Rdata')
```
## Honorees

- Number of honorees up until 2019: `r nrow(keynotes)`.

- Three types of honorees:
```{r}
keynotes %>% 
  pull(conference) %>% 
  table()
```

There are `r nrow(keynotes)` entries and `r keynotes %>% distinct(fore_name, last_name) %>% nrow()` unique names in the ISCB cohort.

Names sorted by the number of honors:
```{r}
keynotes %>% 
  count(fore_name, last_name) %>% 
  arrange(desc(n)) %>% 
  DT::datatable()
```

Number of keynote speakers/fellows across years:

```{r}
keynotes %>%
  select(year, conference) %>% 
  count(year, conference) %>% 
  ggplot(aes(x = year(year), y = n, color = conference)) +
  geom_point() +
  geom_line(alpha = 0.5) +
  coord_cartesian(ylim = c(0, 13)) +
  scale_x_continuous(breaks = seq(1995, 2019, 5)) +
  scale_y_continuous(breaks = seq(0, 14, 2)) +
  scale_color_viridis_d() +
  labs(x = 'Year', y = 'Number of keynote speakers/fellows')
```


## Authors

10 journals with the most computational biology articles:

```{r}
corr_authors %>% 
  mutate(publication_date = ymd(publication_date, truncated = 2)) %>% 
  select(publication_date, journal) %>% 
  filter(journal %in% large_jours$journal) %>% 
  ggplot(aes(x = publication_date, fill = forcats::fct_infreq(journal))) +
  geom_histogram(size = 0, bins = 27) +
  scale_fill_brewer(palette = 'Set3') +
  theme(legend.position = c(0.2, 0.7)) +
  labs(x = NULL, y = NULL) +
scale_x_date(
    labels = scales::date_format("%Y"),
    breaks = as.Date(c('2000-01-01', '2010-01-01', '2019-01-01')),
    limits = c(as.Date('1993-01-01'), as.Date('2019-12-31'))) + 
  # geom_text(data = jours, aes(x = x, y = y, label = label), color = 'grey10') +
  NULL
```

```{r}
nucleic_acids <- corr_authors %>% filter(grepl('Nucleic Acids Res', journal))
table(nucleic_acids$journal)
```

### Gender analysis

```{r}
gender_df <- read_tsv('data/gender/genderize.tsv')

pubmed_gender_pmids <- corr_authors %>%
  left_join(gender_df, by = 'fore_name_simple')

iscb_gender_df <- keynotes %>% 
  left_join(gender_df, by = 'fore_name_simple')
```

`r pubmed_gender_pmids %>% filter(is.na(fore_name)) %>% nrow()` corresponding authors with empty fore name field (i.e., missing metadata).

```{r}
pubmed_gender_pmids %>% 
  mutate(genderized = ifelse(is.na(probability_male), 'NOT genderized', 'Genderized')) %>% 
  ggplot(aes(year(year), fill = genderized)) +
  geom_bar() +
  scale_fill_viridis_d() +
  theme(legend.position = c(0.2, 0.8)) +
  ylab('Number of full names')
```

  
```{r}
gender_check <- pubmed_gender_pmids %>%
  mutate(got_gender = case_when(
    is.na(probability_male) ~ 'Gender not predicted',
    TRUE ~ 'Gender predicted'))

pred_all <- pubmed_gender_pmids %>% 
  count(is.na(probability_male))

dash_df <- pubmed_gender_pmids %>% 
  filter(is.na(probability_male) & !is.na(fore_name_simple)) %>% 
  mutate(dashed_name = grepl('-', fore_name_simple)) %>% 
  count(dashed_name)

pred_before_2002 <- gender_check %>% 
  filter(year(year) < 2002) %>%
  count(is.na(probability_male), is.na(fore_name_simple))

```

In total, `r pred_all[1, 'n']` authors had gender prediction and `r pred_all[2, 'n']` didn't.
`r pubmed_gender_pmids %>% filter(!is.na(fore_name)&is.na(fore_name_simple)) %>% nrow()` authors with fore name that is NA once simplified (i.e., initials only).
Among `r sum(dash_df$n)` authors with fore name but no predictions, ~ `r round(dash_df$n[1]/sum(dash_df$n)*100)`% has a dash.

Honorees that didn't receive a gender prediction: `r iscb_gender_df %>% filter(is.na(probability_male)) %>% pull(full_name.x)`.

Before 2002, `r pred_before_2002[1, 'n']` authors had gender predictions, `r pred_before_2002[2, 'n']` didn't have gender predictions because of these authors only have initials for fore names.


Mean probability of selecting Asian among these names:
```{r}
pubmed_gender_pmids %>% 
  filter(is.na(probability_male) & !is.na(fore_name_simple)) %>% 
  rename('surname' = last_name_simple) %>% 
  predict_race(surname.only = T, impute.missing = F) %>% 
  pull(pred.asi) %>% 
  mean(na.rm = T)
```

**In summary, the NA predictions mostly include initials only, hyphenated names and perhaps names with accent marks.**


```{r include = F, eval = F}
# How much did ISCB honorees publish in these journals as last authors?

n_pubs <- keynotes %>% 
  distinct(fore_name_simple, last_name_simple) %>% 
  # left_join(nat_to_reg, by = c('afflcountries' = 'country_name')) %>% 
  left_join(corr_authors, by = c('fore_name_simple', 'last_name_simple')) %>% 
  group_by(fore_name_simple, last_name_simple) %>% 
  summarise(num_pubs = sum(!is.na(pmid)), .groups = 'drop')

n_pubs %>% 
  ggplot() +
  geom_bar(aes(x = num_pubs))
```



### US race analysis

```{r warning=FALSE, message=FALSE}
pubmed_aff_pmids <- corr_authors %>% 
  tidyr::separate_rows(countries, sep = ',') %>% 
  filter(countries == 'US') # looking at only US affiliation

keynotes_us <- keynotes %>% 
  tidyr::separate_rows(afflcountries, sep = '\\|') %>% 
  filter(afflcountries == 'United States') 

pubmed_race_pmids <- pubmed_aff_pmids %>%
  rename('surname' = last_name_simple) %>% 
  predict_race(surname.only = T, impute.missing = F)

iscb_us_race <- keynotes_us %>% 
  rename('surname' = last_name_simple) %>%
  predict_race(surname.only = T, impute.missing = F)
  
```

- Number of honorees affiliated with the US: `r nrow(keynotes_us)`.

- Number of last authors affiliated with the US: `r nrow(pubmed_aff_pmids)`.

- Number of authors with no race prediction: `r sum(is.na(pubmed_race_pmids$pred.whi))` (of which `r sum(is.na(pubmed_race_pmids$surname))` did not have a surname).

- Number of honorees with no race prediction: `r sum(is.na(iscb_us_race$pred.whi))` (of which `r sum(is.na(iscb_us_race$surname))` did not have a surname).

```{r}
pubmed_race_pmids %>% 
  mutate(race_pred = ifelse(is.na(pred.whi), 'Race NOT predicted', 'Race predicted')) %>% 
  ggplot(aes(year(year), fill = fct_rev(race_pred))) +
  geom_bar() +
  scale_fill_viridis_d() +
  theme(legend.position = c(0.2, 0.8)) +
  ylab('Number of full names')
```


## Name origin analysis

```{r}
region_levels <- paste(c('Celtic/English', 'European', 'East Asian', 'Hispanic', 'South Asian', 'Arabic', 'Hebrew', 'African', 'Nordic', 'Greek'), 'names')

pubmed_nat_pmids <- corr_authors %>%
  left_join(nationalize_df, by = c('fore_name', 'last_name'))

pubmed_nat_df <- pubmed_nat_pmids %>% 
  group_by(pmid, journal, publication_date, year) %>%
  summarise_at(vars(African:SouthAsian), mean, na.rm = T) %>%
  ungroup()

iscb_nat_df <- keynotes %>%
  left_join(nationalize_df, by = c('fore_name', 'last_name'))

```

```{r}
set.seed(0)

top_names <- map_dfc(
  nationalize_df %>% 
    select(African:SouthAsian) %>% 
    colnames(), 
  function(x) {
    nationalize_df %>%
      filter((!!sym(x)) > 0.9) %>%
      sample_n(6) %>%
      select(full_name) %>%
      rename(!!x := full_name)
  })

top_names %>% 
  pivot_longer(everything(), names_to = 'region', values_to = 'names') %>% 
  recode_region() %>% 
  group_by(region) %>% 
  summarise(names = paste(names, collapse = ', '), .groups = 'drop') %>% 
  arrange(factor(region, levels = region_levels)) %>%
  DT::datatable() %>% 
  # write_tsv('data/names/example_name_origin.tsv') %>% 
  {.}
```

`r iscb_nat_df %>% filter(is.na(African)) %>% nrow()` ISCB speakers did not have nationality predictions.
`r pubmed_nat_pmids %>% filter(is.na(African)) %>% nrow()` pubmed full names not nationalized.
`r pubmed_nat_pmids %>% filter(is.na(African)&is.na(fore_name_simple.x)) %>% nrow()` of these don't have fore_name_simple.
(See earlier conclusion in Gender analysis.)

```{r}
pubmed_nat_df %>% 
  mutate(nationalized = ifelse(is.na(African), 'NOT nationalized', 'Nationalized')) %>% 
  ggplot(aes(year(year), fill = nationalized)) +
  geom_bar() +
  scale_fill_viridis_d() +
  theme(legend.position = c(0.2, 0.8)) +
  ylab('Number of full names')
```


## US name origin analysis


```{r}
pubmed_nat_pmids <- corr_authors %>%
  separate_rows(countries, sep = ',') %>% 
  filter(countries == 'US') %>% 
  left_join(nationalize_df, by = c('fore_name', 'last_name'))

pubmed_nat_df <- pubmed_nat_pmids %>% 
  group_by(pmid, journal, publication_date, year) %>%
  summarise_at(vars(African:SouthAsian), mean, na.rm = T) %>%
  ungroup()

iscb_nat_df <- keynotes %>%
  separate_rows(afflcountries, sep = '\\|') %>% 
  filter(afflcountries == 'United States') %>% 
  left_join(nationalize_df, by = c('fore_name', 'last_name'))
```

`r iscb_nat_df %>% filter(is.na(African)) %>% nrow()` US-affiliated ISCB speakers did not have nationality predictions.
`pubmed_nat_pmids %>% filter(is.na(African)) %>% nrow()` US-affiliated authors did not have nationality predictions.
`r pubmed_nat_pmids %>% filter(is.na(African)&is.na(fore_name_simple.x)) %>% nrow()` of these don't have fore_name_simple.
(See earlier conclusion in Gender analysis.)


```{r}
pubmed_nat_df %>% 
  mutate(nationalized = ifelse(is.na(African), 'NOT nationalized', 'Nationalized')) %>% 
  ggplot(aes(year(year), fill = nationalized)) +
  geom_bar() +
  scale_fill_viridis_d() +
  theme(legend.position = c(0.2, 0.8)) +
  ylab('Number of US-affiliated full names')
```




