---
title: "Predict race/ethnicity based on census data"
output: html_document
---

## Setups

```{r message = F, warning=F}
if (!requireNamespace('wru')) install.packages('wru')

library(wru)
library(tidyverse)
library(lubridate)
source('utils/r-utils.R')

theme_set(theme_bw() + theme(legend.title = element_blank()))
```

## Load data
and use the `wru` package to predict race based on last_names only

```{r}
all_full_names <- read_tsv('data/names/full-names.tsv.xz')
articles <- read_tsv('data/pubmed/articles.tsv.xz')

pubmed_race_pmids <- 
  read_tsv('data/names/corresponding-authors.tsv.xz') %>%
  left_join(read_tsv('data/pubmed/articles.tsv.xz'), by = 'pmid') %>%
  mutate(year = substr(publication_date, 1, 4) %>% ymd(truncated = 2),
         publication_date = ymd(publication_date, truncated = 2)) %>%
  rename('surname' = last_name_simple) %>% 
  predict_race(surname.only = T, impute.missing = F)

pubmed_race_df <- pubmed_race_pmids %>% 
  group_by(pmid, journal, publication_date, year) %>% 
  summarise_at(vars(pred.whi:pred.oth), mean, na.rm = T) %>% 
  ungroup()

iscb_race_df <- read_tsv('data/iscb/keynotes.tsv') %>%
  left_join(all_full_names, by = c('fore_name', 'last_name')) %>% 
  rename('surname' = last_name_simple) %>%
  mutate(year = ymd(year, truncated = 2)) %>% 
  predict_race(surname.only = T, impute.missing = F)

alpha_threshold <- qnorm(0.975)
start_year <- 1997
end_year <- 2019
n_years <- end_year - start_year
my_jours <- unique(pubmed_race_df$journal)
my_confs <- unique(iscb_race_df$conference)
n_jours <- length(my_jours)
n_confs <- length(my_confs)
race_levels <- c('White', 'Asian', 'Black', 'Hispanic', 'Other', 'Other categories')
```

Missing data

```{r}
# 10 random unmatched names
iscb_race_df %>% 
  filter(is.na(pred.whi)) %>% 
  select(1,2) %>% 
  sample_n(10)

pubmed_race_pmids %>% 
  filter(is.na(last_name)) %>% 
  nrow()

pubmed_race_pmids %>% 
  filter(is.na(surname)) %>% 
  nrow()

pubmed_race_pmids %>% 
  filter(!is.na(surname)&is.na(pred.whi)) %>% 
  nrow()

iscb_race_df %>% 
  filter(is.na(pred.whi)) %>% 
  nrow()

iscb_race_df %>% 
  filter(is.na(pred.whi)) %>% 
  select(surname) %>%
  sample_n(10)
```

Pubmed authors: 7188 articles out of 29749 did not have race/ethnicity predictions.
8770 names out of 34,010 names did not have race/ethnicity predictions.
40 corresponding authors with empty last name field.
ISCB honorees: 98 names out of 411 did not have race/ethnicity predictions.

## Prepare data frames for later analyses

- bind_rows results of race predictions in iscb and Pubmed authors
- pivot long
- compute mean, sd, marginal error
- add "Other categories"
- separate by journals
- separate by conference keynote speakers/fellows

```{r}
iscb_pubmed <- iscb_race_df %>%
  rename('journal' = conference) %>% 
  select(year, journal, contains('pred')) %>%
  mutate(publication_date = year,
         type = 'Keynote speakers/Fellows') %>%
  bind_rows(
    pubmed_race_df %>%
      select(year, journal, contains('pred'), publication_date) %>%
      mutate(type = 'Pubmed authors')
  ) %>%
  mutate(pred_sum_others = pred.his + pred.oth + pred.bla) %>% 
  pivot_longer(contains('pred'),
               names_to = 'Race',
               values_to = 'probabilities') %>%
  recode_race() %>%
  group_by(type, year, publication_date, Race) %>%
  add_count() %>%
  mutate(
    mean_prob = mean(probabilities, na.rm = T),
    sd_prob = sd(probabilities, na.rm = T),
    n = mean(n),
    me_prob = alpha_threshold * sd_prob / sqrt(n)
  ) %>%
  ungroup() 

```

### By journals

```{r}
pubmed_race <- vector('list', length = n_jours)
i <- 0

for (jour in my_jours){
  i <- i + 1
  pubmed_race[[i]] <- iscb_pubmed %>%
    filter(type == 'Pubmed authors' & journal == jour & (Race != 'Other categories')) %>%
    group_by(year, Race, journal) %>%
    summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
    ungroup()
}
```

### By conference keynote speakers/fellows

```{r}
iscb_race <- vector('list', length = n_confs)
i <- 0

for (conf in my_confs){
  i <- i + 1
  iscb_race[[i]] <- iscb_pubmed %>%
    filter(type != 'Pubmed authors' & journal == conf & (Race != 'Other categories')) %>%
    group_by(year, Race, journal) %>%
    summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
    ungroup()
}
```

```{r include = F, eval = F}
# Here, each circle is a name mention.
# The vertical intervals are confidence intervals of the mean probability of selecting a race/ethnicity category in the US census (triangle).

ggplot(iscb_pubmed %>% filter(Race == 'White' | Race == 'Asian' | Race == 'Other categories'),
       aes(x = year)) +
  facet_grid(cols = vars(fct_rev(Race))) +
  geom_smooth(aes(x = publication_date, y = probabilities, color = type, fill = type)) +
  geom_point(
    data = . %>% filter(type != 'Pubmed authors'),
    aes(y = probabilities, color = type),
    alpha = 0.2,
    show.legend = F
  ) +
  geom_point(
    data = . %>% get_keynote_summary(),
    aes(y = mean_prob, color = type),
    shape = 2,
    alpha = 0.5,
    show.legend = F
  ) +
  geom_linerange(
    data = . %>% get_keynote_summary(),
    aes(
      ymin = mean_prob - me_prob,
      ymax = mean_prob + me_prob,
      color = type
    ),
    alpha = 0.5,
    show.legend = F
  ) +
  scale_y_continuous(breaks = seq(0, 1, 0.2)) +
  scale_x_date(breaks = '4 years', labels = scales::date_format("%Y")) +
  scale_color_viridis_d(option = 'D', end = 0.9) +
  scale_fill_viridis_d(option = 'D', end = 0.9) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = NULL)

```



```{r}
# Check: probabilities, if valid, should add up to 1
iscb_pubmed %>%
  filter(Race == 'White' | Race == 'Asian' | Race == 'Other categories') %>% 
  group_by(type, year, publication_date, Race) %>%
  summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
  summarise(all_prob = sum(mean_prob))
```



## Figures for paper

### Figure 2: Overrepresentation of white and underrepresentation of Asian honorees as compared to authors.

```{r fig.height=8, fig.width=8, warning=FALSE, message=FALSE}
# A simpler figure
fig_stats <- iscb_pubmed %>% 
  filter(Race %in% c('White', 'Asian', 'Other categories')) %>% 
  loess_and_ci() +
  theme(legend.position = c(0.84, 0.78)) +
  facet_wrap(vars(fct_rev(Race)))

fig_2a <- iscb_pubmed %>%
  filter(Race != 'Other categories',
         type == 'Pubmed authors') %>%
  group_by(year, type, Race) %>%
  summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
  ungroup() %>%
  race_breakdown(category = 'main', type) 

fig_2c <- iscb_pubmed %>%
  filter(Race != 'Other categories',
         type != 'Pubmed authors') %>%
  group_by(year, type, Race) %>%
  summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
  ungroup() %>%
  race_breakdown(category = 'main', type)

fig_2b <- bind_rows(pubmed_race) %>%
  race_breakdown(category = 'sub', journal) +
  theme(legend.position = c(0.8, 0.25),
        legend.key.height = unit(4, 'mm'),
        legend.background = element_rect(fill = "transparent", colour = "transparent"))

fig_2d <- bind_rows(iscb_race) %>%
  race_breakdown(category = 'sub', journal)

fig_2_a_d <- cowplot::plot_grid(fig_2a, fig_2b, fig_2c, fig_2d, labels = 'AUTO')
fig_2 <- cowplot::plot_grid(fig_2_a_d, fig_stats, labels = c('', 'E'), ncol = 1, rel_heights = c(2,1))
fig_2

ggsave('figs/racial_makeup.png', fig_2, width = 7, height = 7)
```

### Supplementary Figure S2 {#sup_fig_s2}

For each race/ethnicity category, the mean predicted probability of Pubmed articles is shown as teal LOESS curve, and the mean probability and 95% confidence interval of the ISCB honoree predictions are shown as dark circles and vertical lines.

```{r message=F, warning=F}
fig_s2 <- iscb_pubmed %>% 
  filter(Race != 'Other categories') %>% 
  loess_and_ci() +
  theme(legend.position = c(0.84, 0.3)) +
  facet_wrap(vars(fct_relevel(Race, race_levels)))
fig_s2
ggsave('figs/fig_s2.png', fig_s2, width = 6, height = 4)
```

All Asian speakers at PSB 2000?
```{r}
iscb_race_df %>% filter(year == '2000-01-01', conference == 'PSB')
```

PSB only had one keynote speaker in 2000, Wah Chiu.


What's in Bioinformatics 1999?

```{r}
pubmed_race_pmids %>% 
  filter(year == '1999-01-01' & !is.na(pred.whi)) %>% 
  select(contains('pred')) %>% 
  colMeans() 

pubmed_race_pmids %>% 
  filter(year == '1999-01-01' & !is.na(pred.whi)) %>% 
  filter(pred.asi > 0.1) %>% 
  select(fore_name, surname, pred.asi) 

pubmed_race_pmids %>% 
  filter(year == '1999-01-01' & !is.na(pred.whi)) %>% 
  filter(pred.his > 0.1) %>% 
  select(fore_name, surname, pred.his) 
```

Bioinformatics 1998: `wru` was able to predict 77 out of 121 last authors published that year, and it predicts 10 and 6 authors to have > 0.1 probability of being Asian and Hispanic, respectively, hence what we see.

<!-- ### Statistics? -->
<!-- Confidence interval of the mean? (Aggregation) -->

<!-- Null: The proportion of white keynote iscb speakers is not larger the proportions of white last authors published in Pubmed authors. -->
<!-- Alternative: The proportion of white keynote iscb speakers is larger the proportions of white last authors published in Pubmed authors. -->
<!-- How do we take time component into account? -->
<!-- Right now, I'm thinking we can use `year` as a random variable. -->
<!-- Perhaps we should allow for random slope as well? -->

```{r}
sessionInfo()
```

