---
title: "Visualize nationality predictions"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(rnaturalearth)
source('utils/r-utils.R')
theme_set(theme_bw() + theme(legend.title = element_blank()))
```


```{r}
all_full_names <- read_tsv('data/names/full-names.tsv.xz') %>% distinct()

nationalize_df <- read_tsv('https://raw.githubusercontent.com/greenelab/wiki-nationality-estimate/7c22d0a5f661ce5aeb785215095deda40973ff17/data/NamePrism_results_authors.tsv') %>%
  rename('full_name' = X1) %>%
  distinct(full_name, .keep_all = T) %>%
  left_join(all_full_names, by = 'full_name')

alpha_threshold <- qnorm(0.975)

pubmed_nat_pmids <- read_tsv('data/names/corresponding-authors.tsv.xz') %>%
  left_join(read_tsv('data/pubmed/articles.tsv.xz'), by = 'pmid') %>%
  mutate(year = substr(publication_date, 1, 4) %>% ymd(truncated = 2),
         publication_date = ymd(publication_date, truncated = 2)) %>%
  left_join(nationalize_df, by = c('fore_name', 'last_name'))

pubmed_nat_df <- pubmed_nat_pmids %>%
  group_by(pmid, journal, publication_date, year) %>%
  summarise_at(vars(African:SouthAsian), mean, na.rm = T) %>%
  ungroup()

iscb_nat_df <- read_tsv('data/iscb/keynotes.tsv') %>%
  mutate(publication_date = ymd(year, truncated = 2),
         year = ymd(year, truncated = 2)) %>%
  left_join(nationalize_df, by = c('fore_name', 'last_name')) %>% 
  filter(conference != 'PSB') # remove PSB

start_year <- 1992
end_year <- 2019
n_years <- end_year - start_year
my_jours <- unique(pubmed_nat_df$journal)
my_confs <- unique(iscb_nat_df$conference)
n_jours <- length(my_jours)
n_confs <- length(my_confs)
region_levels <- paste(c('Celtic/English', 'European', 'East Asian', 'Hispanic', 'South Asian', 'Arabic', 'Hebrew', 'African', 'Nordic', 'Greek'), 'names')
region_levels_let <- toupper(letters[1:8])
region_cols <- c('#ffffb3', '#fccde5', '#b3de69', '#fdb462', '#80b1d3', '#8dd3c7', '#bebada', '#fb8072', '#bc80bd', '#ccebc5')
```

Names grouping:
```{r fig.height = 3}
our_country_map <- read_tsv('https://raw.githubusercontent.com/greenelab/wiki-nationality-estimate/7c22d0a5f661ce5aeb785215095deda40973ff17/data/country_to_region_NamePrism.tsv') %>%
  rename('region' = Region) %>%
  # recode_region_letter()
  recode_region()

my_world <- ne_countries(scale='medium', returnclass = 'sf') %>%
  rename(Country = 'name') %>%
  left_join(our_country_map, by = 'Country')

(gworld <- ggplot(data = my_world) +
    geom_sf(aes(fill = fct_relevel(region, region_levels))) +
    coord_sf(crs = "+proj=eqearth +wktext") +
    scale_fill_manual(values = region_cols,
                      na.translate = FALSE) +
    theme(panel.background = element_rect(fill = "azure"),
          legend.title = element_blank(),
          legend.position = 'bottom',
          panel.border = element_rect(fill = NA)))

ggsave('figs/2020-01-31_groupings.png', gworld, width = 7, height = 4)
```

```{r}
set.seed(0)
top_names <- map_dfc(colnames(nationalize_df)[-1], function(x) {
  nationalize_df %>%
    filter((!!sym(x)) > 0.9) %>%
    sample_n(10) %>%
    select(full_name) %>%
    rename(!!x := full_name)
  })

top_names %>% write_tsv('data/names/example_name_origin.tsv')
```


## Missing data

```{r}
iscb_nat_df %>% filter(is.na(African))
pubmed_nat_pmids %>% filter(is.na(African))
pubmed_nat_pmids %>% filter(is.na(African)&!is.na(fore_name_simple.x))
sum(is.na(pubmed_nat_df$African))
```

All ISCB speakers have nationality predictions.
1506 pubmed full names not nationalized because they don't have fore_name_simple.
<!-- These mostly include initials only, hyphenated names and perhaps names with accent marks. -->

```{r}
pubmed_nat_df %>% filter(is.na(African)) %>%
  count(year, journal) %>%
  ggplot(aes(year(year), n, fill = journal)) +
  geom_bar(stat = 'identity') +
  scale_fill_viridis_d() +
  ylab('Number of full names NOT nationalized')
```

In 2000, there is only one article that has nationality prediction.
(9 articles in 2001).

## Descriptive statistics
Prepare data frames for later analyses:

- rbind results of race predictions in iscb and Pubmed
- pivot long
- compute mean, sd, marginal error

```{r}
iscb_pubmed_oth <- iscb_nat_df %>%
  rename('journal' = conference) %>%
  select(year, journal, African:SouthAsian, publication_date) %>%
  mutate(type = 'Keynote speakers/Fellows') %>%
  bind_rows(
    pubmed_nat_df %>%
      select(year, journal, African:SouthAsian, publication_date) %>%
      mutate(type = 'Pubmed authors')
  ) %>%
  mutate(OtherCategories = SouthAsian + Hispanic + Jewish + Muslim + Nordic + Greek + African) %>%
  pivot_longer(c(African:SouthAsian, OtherCategories),
               names_to = 'region',
               values_to = 'probabilities') %>%
  group_by(type, year, publication_date, region) %>%
  add_count() %>%
  mutate(
    mean_prob = mean(probabilities, na.rm = T),
    sd_prob = sd(probabilities, na.rm = T),
    n = mean(n),
    me_prob = alpha_threshold * sd_prob / sqrt(n)
  ) %>%
  ungroup()

iscb_pubmed <- iscb_pubmed_oth %>%
  filter(region != 'OtherCategories')
```

## Prepare data frames for analysis
### By journals

```{r}
pubmed_nat <- vector('list', length = n_jours)
i <- 0

for (jour in my_jours){
  i <- i + 1
  pubmed_nat[[i]] <- iscb_pubmed %>%
    filter(type == 'Pubmed authors' & journal == jour) %>%
    group_by(year, region, journal) %>%
    summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
    ungroup()
}
```

### By conference keynotes/fellows
```{r fig.height=6}
i <- 0
iscb_nat <- vector('list', length = n_confs)

for (conf in my_confs){
  i <- i + 1
  iscb_nat[[i]] <- iscb_pubmed %>%
    filter(type != 'Pubmed authors' & journal == conf) %>%
    group_by(year, region, journal) %>%
    summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
    ungroup()
}
```

## Figures for paper

### Figure 4: Compared to the name collection of Pubmed authors, honorees with Celtic/English names are overrepresented while honorees with East Asian names are underrepresented.

```{r fig.height=7, fig.width=9}
fig_4a <- iscb_pubmed %>%
  filter(year < '2020-01-01') %>%
  group_by(year, type, region) %>%
  summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
  ungroup() %>%
  region_breakdown('main', fct_rev(type)) +
  guides(fill = guide_legend(nrow = 2)) +
  theme(legend.position = 'bottom',
        legend.key.height = unit(3, 'mm'),
        legend.key.width = unit(3, 'mm'),
        legend.text = element_text(size = 8), 
        legend.margin = margin(-0.2, 0.2, 0.2, 0, unit='cm'))

large_regions <- c('CelticEnglish', 'EastAsian', 'European', 'OtherCategories')
## Mean and standard deviation of predicted probabilities:
fig_4b <- iscb_pubmed_oth %>%
  filter(region %in% large_regions) %>%
  # recode_region_letter() %>%
  recode_region() %>%
  loess_and_ci() +
  theme(legend.position = c(0.88, 0.83),
        legend.key.height = unit(4, 'mm'),
        legend.key.width = unit(5, 'mm'),
        legend.text = element_text(size = 6)) +
  facet_wrap(vars(fct_relevel(region, large_regions)), nrow = 1)

fig_4 <- cowplot::plot_grid(fig_4a, fig_4b, labels = 'AUTO', ncol = 1, rel_heights = c(1.3,1))
fig_4
ggsave('figs/region_breakdown.png', fig_4, width = 6.5, height = 5.5)

```


### What about p-values?

```{r}
w_df <- iscb_pubmed %>% filter(region == 'CelticEnglish', !is.na(probabilities))
w_pub <- w_df %>% filter(type == 'Pubmed authors') %>% pull(probabilities)
w_iscb <- w_df %>% filter(type != 'Pubmed authors') %>% pull(probabilities)

t.test(w_iscb, w_pub, alternative = 'greater') %>% 
  broom::tidy() %>% 
  select(-contains('estimate')) %>% 
  print()
```
The Welch two-sample one-sided t-test shows that the mean probability of ISCB speakers predicted to be in group A is higher than that of authors.

```{r}
w_df <- iscb_pubmed %>% filter(region == 'European', !is.na(probabilities))
w_pub <- w_df %>% filter(type == 'Pubmed authors') %>% pull(probabilities)
w_iscb <- w_df %>% filter(type != 'Pubmed authors') %>% pull(probabilities)

t.test(w_iscb, w_pub) %>% 
  broom::tidy() %>% 
  select(-contains('estimate')) %>% 
  print()
```

A Welch two-sample t-test also shows a lower mean probability of ISCB speakers predicted to be in group B compared to authors.

```{r}
w_df <- iscb_pubmed %>% filter(region == 'EastAsian', !is.na(probabilities))
w_pub <- w_df %>% filter(type == 'Pubmed authors') %>% pull(probabilities)
w_iscb <- w_df %>% filter(type != 'Pubmed authors') %>% pull(probabilities)

t.test(w_iscb, w_pub, alternative = 'less') %>% 
  broom::tidy() %>% 
  select(-contains('estimate')) %>% 
  print()
```
The Welch two-sample one-sided t-test shows that the mean probability of ISCB speakers predicted to be in group C is lower than that of authors.

```{r}
w_df <- iscb_pubmed_oth %>% filter(region == 'OtherCategories', !is.na(probabilities))
w_pub <- w_df %>% filter(type == 'Pubmed authors') %>% pull(probabilities)
w_iscb <- w_df %>% filter(type != 'Pubmed authors') %>% pull(probabilities)

t.test(w_iscb, w_pub) %>% 
  broom::tidy() %>% 
  select(-contains('estimate')) %>% 
  print()
```

A Welch two-sample t-test does not reveal any difference in mean probability of ISCB speakers predicted to be in other groups compared to authors.

### Supplementary Figure S4 {#sup_fig_s4}

Proportion of authors in the Celtic English categories had decreased, particularly for papers published in Bioinformatics and BMC Bioinformatics.
ISMB keynotes had more probability attributable to Hebrew names, while RECOMB had more attributable to East Asian names.

```{r}
fig_s4_1 <- bind_rows(pubmed_nat) %>%
  region_breakdown(category = 'sub', journal)

fig_s4_2 <- bind_rows(iscb_nat) %>%
  filter(year < '2020-01-01') %>%
  region_breakdown(category = 'sub', journal)

fig_s4 <- cowplot::plot_grid(fig_s4_1, fig_s4_2, labels = 'AUTO', ncol = 1)
fig_s4
ggsave('figs/fig_s4.png', fig_s4, width = 6, height = 3.5)
```

### Supplementary Figure S5 {#sup_fig_s5}
It's difficult to come to a conclusion for other regions with so few data points and the imperfect accuracy of our prediction.
There seems to be little difference between the proportion of keynote speakers of African, Arabic, South Asian and Hispanic origin than those in the field.
However, just because a nationality isn't underrepresented against the field doesn't mean scientists from that nationality are appropriately represented.

```{r fig.height=6}
fig_s5 <- iscb_pubmed %>%
  recode_region() %>%
  loess_and_ci() +
  theme(legend.position = c(0.8, 0.15)) +
  facet_wrap(vars(fct_relevel(region, region_levels_let)), ncol = 3)
fig_s5
ggsave('figs/fig_s5.png', fig_s5, width = 6, height = 6)

```


```{r}
sessionInfo()
```
