---
title: "Visualize nationality predictions"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
source('utils/r-utils.R')
theme_set(theme_bw() + theme(legend.title = element_blank()))

```

```{r}
nationalize_df <- read_tsv('https://raw.githubusercontent.com/greenelab/wiki-nationality-estimate/05d1c44b5cd5b145b19242c55665993085e08910/data/LSTM_results_authors.tsv') %>% 
  distinct(full_name, .keep_all = T)

all_full_names <- read_tsv('data/names/full-names.tsv.xz')

alpha_threshold <- qnorm(0.975)

pubmed_nat_pmids <- read_tsv('data/names/corresponding-authors.tsv.xz') %>%
  left_join(read_tsv('data/pubmed/articles.tsv.xz'), by = 'pmid') %>%
  mutate(year = substr(publication_date, 1, 4) %>% ymd(truncated = 2),
         publication_date = ymd(publication_date, truncated = 2)) %>%
  mutate(full_name = paste(fore_name, last_name)) %>% 
  left_join(nationalize_df, by = 'full_name')

pubmed_nat_df <- pubmed_nat_pmids %>% 
  group_by(pmid, journal, publication_date, year) %>% 
  summarise_at(vars(African:SouthAsian), mean, na.rm = T) %>% 
  ungroup()

iscb_nat_df <- read_tsv('data/iscb/keynotes.tsv') %>%
  mutate(publication_date = ymd(year, truncated = 2),
         year = ymd(year, truncated = 2)) %>% 
  select(- full_name) %>% 
  left_join(all_full_names, by = c('fore_name', 'last_name')) %>% 
  left_join(nationalize_df, by = 'full_name') 

start_year <- 1997
end_year <- 2019
n_years <- end_year - start_year
my_jours <- unique(pubmed_nat_df$journal)
my_confs <- unique(iscb_nat_df$conference)
n_jours <- length(my_jours)
n_confs <- length(my_confs)
region_levels <- c('Celtic English', 'European', 'East Asian', 'Hispanic', 'South Asian', 'Muslim', 'Israel', 'African')
```

```{r}
set.seed(0)
top_names <- map_dfc(colnames(nationalize_df)[-1], function(x) {
  nationalize_df %>% 
    filter((!!sym(x)) > 0.9) %>% 
    sample_n(10) %>% 
    select(full_name) %>% 
    rename(!!x := full_name)
  }) 

top_names %>% write_tsv('data/names/example_nationalities.tsv')
```


## Missing data

```{r}
iscb_nat_df %>% filter(is.na(African)) 
pubmed_nat_pmids %>% filter(is.na(African)) %>% sample_n(10)
sum(is.na(pubmed_nat_df$African))
```

1457 pubmed full names not nationalized
<!-- These mostly include initials only, hyphenated names and perhaps names with accent marks. -->

```{r}
pubmed_nat_df %>% filter(is.na(African)) %>% 
  count(year, journal) %>% 
  ggplot(aes(year(year), n, fill = journal)) +
  geom_bar(stat = 'identity') +
  scale_fill_viridis_d() +
  ylab('Number of full names NOT nationalized')
```


```{r}
pubmed_nat_df %>% 
  filter(year == '2000-01-01', !is.na(African))
```

In 2010, there is only one article that has nationality prediction.
Should we drop?

## Descriptive statistics
Prepare data frames for later analyses:

- rbind results of race predictions in iscb and Pubmed
<!-- - pivot long -->
- compute mean, sd, marginal error


```{r}
iscb_pubmed <- iscb_nat_df %>%
  rename('journal' = conference) %>% 
  select(year, journal, African:SouthAsian, publication_date) %>%
  mutate(type = 'Keynote speakers/Fellows') %>%
  bind_rows(
    pubmed_nat_df %>%
      select(year, journal, African:SouthAsian, publication_date) %>%
      mutate(type = 'Pubmed authors')
  ) %>%
  # mutate(probability_female = 1 - African) %>% 
  pivot_longer(African:SouthAsian,
               names_to = 'region',
               values_to = 'probabilities') %>%
  group_by(type, year, publication_date, region) %>%
  add_count() %>%
  mutate(
    mean_prob = mean(probabilities, na.rm = T),
    sd_prob = sd(probabilities, na.rm = T),
    n = mean(n),
    me_prob = alpha_threshold * sd_prob / sqrt(n)
  ) %>%
  ungroup() 
```

## Prepare data frames for analysis
### By journals

```{r}
pubmed_nat <- vector('list', length = n_jours)
i <- 0

for (jour in my_jours){
  i <- i + 1
  pubmed_nat[[i]] <- iscb_pubmed %>%
    filter(type == 'Pubmed authors' & journal == jour) %>%
    group_by(year, region, journal) %>%
    summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
    ungroup() 
}
```

### By conference keynotes/fellows
```{r fig.height=6}
i <- 0
iscb_nat <- vector('list', length = n_confs)

for (conf in my_confs){
  i <- i + 1
  iscb_nat[[i]] <- iscb_pubmed %>%
    filter(type != 'Pubmed authors' & journal == conf) %>%
    group_by(year, region, journal) %>%
    summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
    ungroup()
}
```


```{r}
iscb_pubmed %>%
    filter(type != 'Pubmed authors' & year == '2011-01-01' & journal == 'PSB')
iscb_nat_df %>% filter(year == '2011-01-01' & conference == 'PSB')
iscb_nat_df %>% filter(year == '2020-01-01' & conference == 'PSB')
```

## Comparisons

```{r}
fig_4a <- iscb_pubmed %>%
  # filter(type == 'Pubmed authors') %>%
  filter(year < '2020-01-01') %>% 
  group_by(year, type, region) %>%
  summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
  ungroup() %>%
  region_breakdown('main', fct_rev(type)) +
  guides(fill = guide_legend(nrow = 1)) +
  theme(legend.position = 'bottom',
        legend.key.height = unit(3, 'mm'),
        legend.key.width = unit(3, 'mm'),
        legend.text = element_text(size = 8),
        legend.margin = margin(-0.2, 0.2, 0.2, 0, unit='cm'))

# fig_4a <- iscb_pubmed %>%
#   filter(type != 'Pubmed authors') %>%
#   group_by(year, type, region) %>%
#   summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
#   ungroup() %>%
#   region_breakdown('main', type)


# fig_4_a_d <- cowplot::plot_grid(fig_4a, fig_4b, fig_4c, fig_4d, labels = 'AUTO')
```

## Mean and standard deviation of predicted probabilities

A simpler figure, supplementary Fig S3?

```{r fig.height=8, fig.width=8}
fig_4b <- iscb_nat_df %>%
  rename('journal' = conference) %>% 
  select(year, journal, African:SouthAsian, publication_date) %>%
  mutate(type = 'Keynote speakers/Fellows') %>%
  bind_rows(
    pubmed_nat_df %>%
      select(year, journal, African:SouthAsian, publication_date) %>%
      mutate(type = 'Pubmed authors')
  ) %>%
  mutate(OtherCategories = SouthAsian + Hispanic + Muslim + Israel + African) %>% 
  select(-c(Hispanic, SouthAsian, African, Israel, Muslim)) %>% 
  pivot_longer(c(CelticEnglish, EastAsian, European, OtherCategories),
               names_to = 'region',
               values_to = 'probabilities') %>%
  group_by(type, year, publication_date, region) %>%
  add_count() %>%
  mutate(
    mean_prob = mean(probabilities, na.rm = T),
    sd_prob = sd(probabilities, na.rm = T),
    n = mean(n),
    me_prob = alpha_threshold * sd_prob / sqrt(n)
  ) %>%
  ungroup() %>% 
  recode_region() %>% 
  ggplot(aes(group = type)) +
  geom_smooth(size = 0.2, data = . %>% filter(type == 'Pubmed authors'),
              aes(x = publication_date, y = probabilities, 
                  fill = type, 
                  color = type)) +
  geom_pointrange(data = . %>% get_keynote_summary(),
                  alpha = 0.75, size = 0.3,
                  aes(x = year, y = mean_prob, shape = type,
                      ymin = mean_prob - me_prob,
                      ymax = mean_prob + me_prob)) +
  scale_y_continuous(breaks = seq(0, 1, 0.2), labels = scales::percent_format()) +
  scale_x_date(labels = scales::date_format("'%y"), breaks = '5 years',
               limits = c(ymd(start_year, truncated = 2L),
          ymd(end_year, truncated = 2L))) +
  scale_color_manual(values = '#3fa392') +
  scale_fill_manual(values = '#3fa392') +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = NULL, y = 'Estimated composition') +
  theme(panel.grid.minor = element_blank(),
        legend.position = c(0.88, 0.83),
        legend.key.height = unit(4, 'mm'),
        legend.key.width = unit(5, 'mm'),
        legend.text = element_text(size = 6),
        legend.margin = margin(-0.5, 0, 0, 0, unit='cm')) +
  facet_wrap(vars(fct_relevel(region, c('Celtic English', 'European', 'East Asian', 'Other categories'))), nrow = 1)

# fig_4 <- cowplot::plot_grid(fig_4a, fig_4b, labels = 'AUTO', ncol = 1, rel_height = c(2, 1))
# fig_4

fig_4 <- cowplot::plot_grid(fig_4a, fig_4b, labels = 'AUTO', ncol = 1, rel_heights = c(1.3,1))

ggsave('figs/region_breakdown.png', fig_4, width = 6.5, height = 5.5)

```

```{r}

fig_s4a <- bind_rows(pubmed_nat) %>%
  region_breakdown(category = 'sub', journal) +
  theme(legend.position = 'right'
        ,
        legend.key.height = unit(3.5, 'mm'),
        legend.margin = margin(0,0.3,0, 0.2, unit='cm')
        # legend.text = element_text(size = 7),
        # legend.background = element_rect(fill = "transparent", colour = "transparent")
        )

fig_s4b <- bind_rows(iscb_nat) %>%
  filter(year < '2020-01-01') %>% 
  region_breakdown(category = 'sub', journal)

fig_s4 <- cowplot::plot_grid(fig_s4a, fig_s4b, labels = 'AUTO', ncol = 1)
ggsave('figs/fig_s4.png', fig_s4, width = 6, height = 3.5)
```

Supplementary figure?
```{r}
fig_s4b <- iscb_pubmed %>%
  ggplot(aes(group = type)) +
  geom_smooth(size = 0.2, data = . %>% filter(type == 'Pubmed authors'),
              aes(x = publication_date, y = probabilities, 
                  fill = type, 
                  color = type)) +
  geom_pointrange(data = . %>% get_keynote_summary(),
                  alpha = 0.75, size = 0.3,
                  aes(x = year, y = mean_prob, shape = type,
                      ymin = mean_prob - me_prob,
                      ymax = mean_prob + me_prob)) +
  scale_y_continuous(breaks = seq(0, 1, 0.2), labels = scales::percent_format()) +
  scale_x_date(labels = scales::date_format("'%y"), breaks = '5 years',
               limits = c(ymd(start_year, truncated = 2L),
          ymd(end_year, truncated = 2L))) +
  scale_color_manual(values = '#8dd3c7') +
  scale_fill_manual(values = '#8dd3c7') +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = NULL, y = 'Estimated composition') +
  theme(panel.grid.minor = element_blank(),
        legend.position = c(0.84, 0.2),
        legend.key.height = unit(4, 'mm'),
        legend.key.width = unit(5, 'mm'),
        legend.text = element_text(size = 8),
        legend.margin = margin(-0.5, 0, 0, 0, unit='cm')) +
  facet_wrap(vars(fct_relevel(region, c('CelticEnglish', 'European', 'EastAsian', 'OtherCategories')))) 

ggsave('figs/fig_s4b.png', fig_s4b, width = 6, height = 6)

```


```{r}
sessionInfo()
```

