---
title: "Name origin analysis in the US"
output:
  html_document:
    theme: flatly
    code_download: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
source('utils/r-utils.R')
theme_set(theme_bw() + theme(legend.title = element_blank()))
```

```{r}
all_full_names <- read_tsv('data/names/full-names.tsv.xz') %>% distinct()

nationalize_df <- read_tsv('https://raw.githubusercontent.com/greenelab/wiki-nationality-estimate/c9e459ab7ee869cef777d0946a25a17f2328932b/data/NamePrism_results_authors.tsv') %>%
  rename('full_name' = X1) %>%
  distinct(full_name, .keep_all = T) %>%
  left_join(all_full_names, by = 'full_name')

alpha_threshold <- qnorm(0.975)

pubmed_nat_pmids <- read_tsv('data/names/corresponding-authors.tsv.xz',
                             col_types = readr::cols(fore_name_simple = readr::col_character())) %>%
  left_join(read_tsv('data/pubmed/articles.tsv.xz'), by = 'pmid') %>%
  mutate(year = substr(publication_date, 1, 4) %>% ymd(truncated = 2),
         publication_date = ymd(publication_date, truncated = 2)) %>%
  separate_rows(countries, sep = ',') %>% 
  filter(countries == 'US') %>% 
  left_join(nationalize_df, by = c('fore_name', 'last_name'))

pubmed_nat_df <- pubmed_nat_pmids %>%
  group_by(pmid, journal, publication_date, year) %>%
  summarise_at(vars(African:SouthAsian), mean, na.rm = T) %>%
  ungroup()

iscb_nat_df <- read_tsv('data/iscb/keynotes.tsv') %>%
  mutate(publication_date = ymd(year, truncated = 2),
         year = ymd(year, truncated = 2)) %>%
  separate_rows(afflcountries, sep = '\\|') %>% 
  filter(afflcountries == 'United States') %>% 
  left_join(nationalize_df, by = c('fore_name', 'last_name')) %>% 
  filter(conference != 'PSB', year != '2020-01-01') # remove PSB, exclude ISCB Fellows and ISMB speakers in 2020 for now

start_year <- 1992
end_year <- 2019
n_years <- end_year - start_year
my_jours <- unique(pubmed_nat_df$journal)
my_confs <- unique(iscb_nat_df$conference)
n_jours <- length(my_jours)
n_confs <- length(my_confs)
region_levels <- paste(c('Celtic/English', 'European', 'East Asian', 'Hispanic', 'South Asian', 'Arabic', 'Hebrew', 'African', 'Nordic', 'Greek'), 'names')

region_cols <- c('#ffffb3', '#fccde5', '#b3de69', '#fdb462', '#80b1d3', '#8dd3c7', '#bebada', '#fb8072', '#bc80bd', '#ccebc5')
```

## Missing data

```{r}
# iscb_nat_df %>% filter(is.na(African))
pubmed_nat_pmids %>% filter(is.na(African))
pubmed_nat_pmids %>% filter(is.na(African)&!is.na(fore_name_simple.x))
sum(is.na(pubmed_nat_df$African))
```

All ISCB speakers have nationality predictions.
`r sum(is.na(pubmed_nat_df$African))` pubmed full names not nationalized because they don't have fore_name_simple (only initials were recorded).
<!-- These mostly include initials only, hyphenated names and perhaps names with accent marks. -->

```{r}
pubmed_nat_df %>% filter(is.na(African)) %>%
  count(year) %>%
  ggplot(aes(year(year), n)) +
  geom_bar(stat = 'identity') +
  scale_fill_viridis_d() +
  ylab('Number of full names NOT nationalized')
```

## Descriptive statistics
Prepare data frames for later analyses:

- rbind results of race predictions in iscb and Pubmed
- pivot long
- compute mean, sd, marginal error

```{r}
iscb_pubmed_oth <- iscb_nat_df %>%
  rename('journal' = conference) %>%
  select(year, journal, African:SouthAsian, publication_date) %>%
  mutate(type = 'Keynote speakers/Fellows') %>%
  bind_rows(
    pubmed_nat_df %>%
      select(year, journal, African:SouthAsian, publication_date) %>%
      mutate(type = 'Pubmed authors')
  ) %>%
  mutate(OtherCategories = SouthAsian + Hispanic + Jewish + Muslim + Nordic + Greek + African) %>%
  pivot_longer(c(African:SouthAsian, OtherCategories),
               names_to = 'region',
               values_to = 'probabilities') %>%
  filter(!is.na(probabilities)) %>% 
  group_by(type, year, publication_date, region) %>%
  add_count() %>%
  mutate(
    mean_prob = mean(probabilities, na.rm = T),
    sd_prob = sd(probabilities, na.rm = T),
    n = mean(n),
    me_prob = alpha_threshold * sd_prob / sqrt(n)
  ) %>%
  ungroup()

iscb_pubmed <- iscb_pubmed_oth %>%
  filter(region != 'OtherCategories')
```

## Prepare data frames for analysis
### By journals

```{r}
# pubmed_nat <- vector('list', length = n_jours)
# i <- 0
# 
# for (jour in my_jours){
#   i <- i + 1
#   pubmed_nat[[i]] <- iscb_pubmed %>%
#     filter(type == 'Pubmed authors' & journal == jour) %>%
#     group_by(year, region, journal) %>%
#     summarise(mean_prob = mean(probabilities, na.rm = T), .groups = 'drop')
# }
```

### By conference keynotes/fellows
```{r fig.height=6}
i <- 0
iscb_nat <- vector('list', length = n_confs)

for (conf in my_confs){
  i <- i + 1
  iscb_nat[[i]] <- iscb_pubmed %>%
    filter(type != 'Pubmed authors' & journal == conf) %>%
    group_by(year, region, journal) %>%
    summarise(mean_prob = mean(probabilities, na.rm = T), .groups = 'drop')
}
```

## Figures for paper

### Figure

```{r fig.height=7, fig.width=9}
fig_us_name_origina <- iscb_pubmed %>%
  filter(year < '2020-01-01') %>%
  group_by(year, type, region) %>%
  summarise(mean_prob = mean(probabilities, na.rm = T), .groups = 'drop') %>% 
  region_breakdown('main', fct_rev(type)) +
  guides(fill = guide_legend(nrow = 2)) +
  theme(legend.position = 'bottom',
        legend.key.height = unit(3, 'mm'),
        legend.key.width = unit(3, 'mm'),
        legend.text = element_text(size = 8),
        legend.margin = margin(-0.2, 0.2, 0.2, 0, unit='cm'))

large_regions <- c('CelticEnglish', 'EastAsian', 'European', 'OtherCategories')
## Mean and standard deviation of predicted probabilities:
fig_us_name_originb <- iscb_pubmed_oth %>%
  filter(region %in% large_regions) %>%
  recode_region() %>%
  loess_and_ci() +
  theme(legend.position = c(0.88, 0.83),
        legend.key.height = unit(4, 'mm'),
        legend.key.width = unit(5, 'mm'),
        legend.text = element_text(size = 6)) +
  facet_wrap(vars(region), nrow = 1)

fig_us_name_origin <- cowplot::plot_grid(fig_us_name_origina, fig_us_name_originb, labels = 'AUTO', ncol = 1, rel_heights = c(1.3,1))
fig_us_name_origin
ggsave('figs/us_name_origin.png', fig_us_name_origin, width = 6.5, height = 5.5)
ggsave('figs/us_name_origin.svg', fig_us_name_origin, width = 6.5, height = 5.5)
```

### What about p-values?

```{r}
inte_lm <- function(regioni){
  lm(probabilities ~ year + type, 
     data = iscb_pubmed_oth %>% 
       filter(region == regioni, !is.na(probabilities)) %>% 
       mutate(year = scale(year)))
}

inte_list <- lapply(large_regions, inte_lm)
names(inte_list) <- large_regions
lapply(inte_list, summary)
```

```{r echo = F}
get_p <- function(i, colu){
  broom::tidy(inte_list[[i]]) %>% 
    filter(term == 'typePubmed authors') %>% 
    pull(colu) %>% 
    sprintf("%0.5g", .)
}
```

## Conclusion

A name coming from the group of authors has significantly lower probability of being Celtic/English, $\beta_\textrm{Celtic/English} =$ `r get_p(1, 'estimate')` (_P_ = `r get_p(1, 'p.value')`), lower probability of being European, $\beta_\textrm{European} =$ `r get_p(3, 'estimate')` (_P_ = `r get_p(3, 'p.value')`), and higher probability of being East Asian, $\beta_\textrm{East Asian} =$ `r get_p(2, 'estimate')` (_P_ = `r get_p(2, 'p.value')`). The two groups of scientists did not have a significant association with names predicted to be in Other categories (_P_ = `r get_p(4, 'p.value')`).




### Supplementary Figure S6 {#sup_fig_s6}

Proportion of authors in the Celtic English categories had decreased, particularly for papers published in Bioinformatics and BMC Bioinformatics.
ISMB keynotes had more probability attributable to Hebrew names, while RECOMB had more attributable to East Asian names.

```{r}
# fig_s6_1 <- bind_rows(pubmed_nat) %>%
#   region_breakdown(category = 'sub', journal)

fig_s6_2 <- bind_rows(iscb_nat) %>%
  filter(year < '2020-01-01') %>%
  region_breakdown(category = 'sub', journal)

# fig_s6 <- cowplot::plot_grid(fig_s6_1, fig_s6_2, labels = 'AUTO', ncol = 1)
# fig_s6
ggsave('figs/fig_s6.png', fig_s6_2, width = 6, height = 3.5)
ggsave('figs/fig_s6.svg', fig_s6_2, width = 6, height = 3.5)
```

### Supplementary Figure S7 {#sup_fig_s7}
It's difficult to come to a conclusion for other regions with so few data points and the imperfect accuracy of our prediction.
There seems to be little difference between the proportion of keynote speakers of African, Arabic, South Asian and Hispanic origin than those in the field.
However, just because a nationality isn't underrepresented against the field doesn't mean scientists from that nationality are appropriately represented.

```{r fig.height=6}
fig_s7 <- iscb_pubmed %>%
  recode_region() %>%
  loess_and_ci() +
  theme(legend.position = c(0.8, 0.1)) +
  facet_wrap(vars(fct_relevel(region, region_levels)), ncol = 3)
fig_s7
ggsave('figs/fig_s7.png', fig_s7, width = 6, height = 6)
ggsave('figs/fig_s7.svg', fig_s7, width = 6, height = 6)
```


```{r}
sessionInfo()
```
