---
title: "Name origin analysis in the US"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
source('utils/r-utils.R')
theme_set(theme_bw() + theme(legend.title = element_blank()))
```

```{r}
all_full_names <- read_tsv('data/names/full-names.tsv.xz') %>% distinct()

nationalize_df <- read_tsv('https://raw.githubusercontent.com/greenelab/wiki-nationality-estimate/7c22d0a5f661ce5aeb785215095deda40973ff17/data/NamePrism_results_authors.tsv') %>%
  rename('full_name' = X1) %>%
  distinct(full_name, .keep_all = T) %>%
  left_join(all_full_names, by = 'full_name')

alpha_threshold <- qnorm(0.975)

pubmed_nat_pmids <- read_tsv('data/names/corresponding-authors.tsv.xz') %>%
  left_join(read_tsv('data/pubmed/articles.tsv.xz'), by = 'pmid') %>%
  mutate(year = substr(publication_date, 1, 4) %>% ymd(truncated = 2),
         publication_date = ymd(publication_date, truncated = 2)) %>%
  separate_rows(countries, sep = ',') %>% 
  filter(countries == 'US') %>% 
  left_join(nationalize_df, by = c('fore_name', 'last_name'))

pubmed_nat_df <- pubmed_nat_pmids %>%
  group_by(pmid, journal, publication_date, year) %>%
  summarise_at(vars(African:SouthAsian), mean, na.rm = T) %>%
  ungroup()

iscb_nat_df <- read_tsv('data/iscb/keynotes.tsv') %>%
  mutate(publication_date = ymd(year, truncated = 2),
         year = ymd(year, truncated = 2)) %>%
  separate_rows(afflcountries, sep = '\\|') %>% 
  filter(afflcountries == 'United States') %>% 
  left_join(nationalize_df, by = c('fore_name', 'last_name')) %>% 
  filter(conference != 'PSB') # remove PSB

start_year <- 1992
end_year <- 2019
n_years <- end_year - start_year
my_jours <- unique(pubmed_nat_df$journal)
my_confs <- unique(iscb_nat_df$conference)
n_jours <- length(my_jours)
n_confs <- length(my_confs)
region_levels <- paste(c('Celtic/English', 'European', 'East Asian', 'Hispanic', 'South Asian', 'Arabic', 'Hebrew', 'African', 'Nordic', 'Greek'), 'names')

region_cols <- c('#ffffb3', '#fccde5', '#b3de69', '#fdb462', '#80b1d3', '#8dd3c7', '#bebada', '#fb8072', '#bc80bd', '#ccebc5')

```

## Missing data

```{r}
iscb_nat_df %>% filter(is.na(African))
pubmed_nat_pmids %>% filter(is.na(African))
pubmed_nat_pmids %>% filter(is.na(African)&!is.na(fore_name_simple.x))
sum(is.na(pubmed_nat_df$African))
```

All ISCB speakers have nationality predictions.
1506 pubmed full names not nationalized because they don't have fore_name_simple.
<!-- These mostly include initials only, hyphenated names and perhaps names with accent marks. -->

```{r}
pubmed_nat_df %>% filter(is.na(African)) %>%
  count(year, journal) %>%
  ggplot(aes(year(year), n, fill = journal)) +
  geom_bar(stat = 'identity') +
  scale_fill_viridis_d() +
  ylab('Number of full names NOT nationalized')
```


```{r}
pubmed_nat_df %>%
  filter(year == '2000-01-01', !is.na(African))
```

In 2000, there is only one article that has nationality prediction.
(9 articles in 2001).
Should we drop?

## Descriptive statistics
Prepare data frames for later analyses:

- rbind results of race predictions in iscb and Pubmed
- pivot long
- compute mean, sd, marginal error

```{r}
iscb_pubmed_oth <- iscb_nat_df %>%
  rename('journal' = conference) %>%
  select(year, journal, African:SouthAsian, publication_date) %>%
  mutate(type = 'Keynote speakers/Fellows') %>%
  bind_rows(
    pubmed_nat_df %>%
      select(year, journal, African:SouthAsian, publication_date) %>%
      mutate(type = 'Pubmed authors')
  ) %>%
  mutate(OtherCategories = SouthAsian + Hispanic + Jewish + Muslim + Nordic + Greek + African) %>%
  pivot_longer(c(African:SouthAsian, OtherCategories),
               names_to = 'region',
               values_to = 'probabilities') %>%
  group_by(type, year, publication_date, region) %>%
  add_count() %>%
  mutate(
    mean_prob = mean(probabilities, na.rm = T),
    sd_prob = sd(probabilities, na.rm = T),
    n = mean(n),
    me_prob = alpha_threshold * sd_prob / sqrt(n)
  ) %>%
  ungroup()

iscb_pubmed <- iscb_pubmed_oth %>%
  filter(region != 'OtherCategories')
```

## Prepare data frames for analysis
### By journals

```{r}
pubmed_nat <- vector('list', length = n_jours)
i <- 0

for (jour in my_jours){
  i <- i + 1
  pubmed_nat[[i]] <- iscb_pubmed %>%
    filter(type == 'Pubmed authors' & journal == jour) %>%
    group_by(year, region, journal) %>%
    summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
    ungroup()
}
```

### By conference keynotes/fellows
```{r fig.height=6}
i <- 0
iscb_nat <- vector('list', length = n_confs)

for (conf in my_confs){
  i <- i + 1
  iscb_nat[[i]] <- iscb_pubmed %>%
    filter(type != 'Pubmed authors' & journal == conf) %>%
    group_by(year, region, journal) %>%
    summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
    ungroup()
}
```

## Figures for paper

### Figure

```{r fig.height=7, fig.width=9}
fig_us_name_origina <- iscb_pubmed %>%
  filter(year < '2020-01-01') %>%
  group_by(year, type, region) %>%
  summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
  ungroup() %>%
  region_breakdown('main', fct_rev(type)) +
  guides(fill = guide_legend(nrow = 2)) +
  theme(legend.position = 'bottom',
        legend.key.height = unit(3, 'mm'),
        legend.key.width = unit(3, 'mm'),
        legend.text = element_text(size = 8),
        legend.margin = margin(-0.2, 0.2, 0.2, 0, unit='cm'))

large_regions <- c('CelticEnglish', 'EastAsian', 'European', 'OtherCategories')
## Mean and standard deviation of predicted probabilities:
fig_us_name_originb <- iscb_pubmed_oth %>%
  filter(region %in% large_regions) %>%
  recode_region() %>%
  loess_and_ci() +
  theme(legend.position = c(0.88, 0.83),
        legend.key.height = unit(4, 'mm'),
        legend.key.width = unit(5, 'mm'),
        legend.text = element_text(size = 6)) +
  facet_wrap(vars(region), nrow = 1)

fig_us_name_origin <- cowplot::plot_grid(fig_us_name_origina, fig_us_name_originb, labels = 'AUTO', ncol = 1, rel_heights = c(1.3,1))
fig_us_name_origin
ggsave('figs/us_name_origin.png', fig_us_name_origin, width = 6.5, height = 5.5)

```


```{r}
sessionInfo()
```
