---
title: 'Affiliation analysis'
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(rnaturalearth)
source('utils/r-utils.R')
theme_set(theme_bw() + theme(legend.title = element_blank()))
```

```{r}
world <- ne_countries(scale='medium',returnclass = 'sf') 
nat_to_reg <- world %>% 
  as_tibble() %>% 
  select(iso_a2, name, region_wb) %>% 
  rename('countries' = iso_a2,
         'country_name' = name,
         'region' = region_wb)
# region_levels <- unique(nat_to_reg$region) %>% 
#   setdiff('Antarctica')
region_levels <- c('North America', 'Europe & Central Asia', 'East Asia & Pacific', 'Latin America & Caribbean', 'South Asia', 'Middle East & North Africa', 'Sub-Saharan Africa')
```

```{r}
all_full_names <- read_tsv('data/names/full-names.tsv.xz')

alpha_threshold <- qnorm(0.975)

pubmed_aff_pmids <- read_tsv('data/names/corresponding-authors.tsv.xz') %>%
  left_join(read_tsv('data/pubmed/articles.tsv.xz'), by = 'pmid') %>%
  mutate(year = substr(publication_date, 1, 4) %>% ymd(truncated = 2),
         publication_date = ymd(publication_date, truncated = 2)) 

pubmed_aff_df <- pubmed_aff_pmids %>%
  filter(!is.na(countries)) %>% # a paper has two corresponding authors, one does not have affiliations. Should we remove this author and give the other the full weight (1) on the paper?
  add_count(pmid, name = 'num_corr_authors') %>% # number of corresponding authors per pmid
  select(pmid, journal, publication_date, year, countries, fore_name_simple, last_name_simple, num_corr_authors) %>%
  separate_rows(countries) %>%
  add_count(pmid, fore_name_simple, last_name_simple, name = 'num_affls') %>% # number of country affiliations per author
  mutate(probabilities = 1 / num_corr_authors / num_affls) %>%
  left_join(nat_to_reg, by = 'countries') %>%
  filter(!is.na(country_name)) %>%
  mutate(row = row_number()) %>%
  pivot_wider(
    names_from = region,
    values_from = probabilities,
    values_fill = list(probabilities = 0)
  ) %>%
  select(-row) %>% 
  group_by(pmid) %>% 
  mutate_at(region_levels, sum) %>% 
  ungroup() %>% 
  distinct(pmid, .keep_all = T) %>% 
  mutate(`Other Categories` = `South Asia` + `Latin America & Caribbean` + `Middle East & North Africa` + `Sub-Saharan Africa`) %>%
  select(-c(countries, country_name, num_affls)) %>% 
  pivot_longer(c(region_levels, `Other Categories`),
               names_to = 'region',
               values_to = 'probabilities')


iscb_aff_df <- read_tsv('data/iscb/keynotes.tsv') %>%
  separate_rows(afflcountries, sep = '\\|') %>% 
  add_count(year, conference, full_name, name = 'num_affls') %>%
  mutate(probabilities = 1 / num_affls,
         publication_date = ymd(year, truncated = 2),
         year = ymd(year, truncated = 2)) %>%
  left_join(nat_to_reg, by = c('afflcountries' = 'country_name')) %>%
  filter(!is.na(afflcountries)) %>%
  pivot_wider(
    names_from = region,
    values_from = probabilities,
    values_fill = list(probabilities = 0)
  ) %>% 
  group_by(year, conference, full_name) %>% 
  mutate_at(vars(`North America`:`East Asia & Pacific`), sum) %>% 
  ungroup() %>% 
  distinct(year, conference, full_name, .keep_all = T) %>% 
  mutate(`Other Categories` = `Latin America & Caribbean` + `Middle East & North Africa`) %>%
  pivot_longer(c(`North America`:`East Asia & Pacific`, `Other Categories`),
               names_to = 'region',
               values_to = 'probabilities')

start_year <- 1997
end_year <- 2019
n_years <- end_year - start_year
my_jours <- unique(pubmed_aff_df$journal)
my_confs <- unique(iscb_aff_df$conference)
n_jours <- length(my_jours)
n_confs <- length(my_confs)

region_levels_let <- toupper(letters[1:8])
```

## Missing data

```{r}
# iscb_aff_df %>% filter(is.na(countries))
pubmed_aff_pmids %>% filter(is.na(countries))
sum(is.na(pubmed_aff_pmids$countries))
sum(is.na(pubmed_aff_df$countries))

```


## Descriptive statistics
Prepare data frames for later analyses:

- rbind results of race predictions in iscb and Pubmed
- pivot long
- compute mean, sd, marginal error

```{r}
iscb_pubmed_oth <- iscb_aff_df %>%
  rename('journal' = conference) %>%
  select(year, journal, publication_date, probabilities, region) %>%
  mutate(type = 'Keynote speakers/Fellows') %>%
  bind_rows(
    pubmed_aff_df %>%
      select(year, journal, publication_date, probabilities, region) %>%
      mutate(type = 'Pubmed authors')
  ) %>%
  group_by(type, year, publication_date, region) %>%
  add_count() %>%
  mutate(
    mean_prob = mean(probabilities, na.rm = T),
    sd_prob = sd(probabilities, na.rm = T),
    n = mean(n),
    me_prob = alpha_threshold * sd_prob / sqrt(n)
  ) %>%
  ungroup()

iscb_pubmed <- iscb_pubmed_oth %>%
  filter(region != 'Other Categories')
```

## Prepare data frames for analysis
### By journals

```{r}
pubmed_aff <- vector('list', length = n_jours)
i <- 0

for (jour in my_jours){
  i <- i + 1
  pubmed_aff[[i]] <- iscb_pubmed %>%
    filter(type == 'Pubmed authors' & journal == jour) %>%
    group_by(year, region, journal) %>%
    summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
    ungroup()
}
```



### By conference keynotes/fellows
```{r fig.height=6}
i <- 0
iscb_aff <- vector('list', length = n_confs)

for (conf in my_confs){
  i <- i + 1
  iscb_aff[[i]] <- iscb_pubmed %>%
    filter(type != 'Pubmed authors' & journal == conf) %>%
    group_by(year, region, journal) %>%
    summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
    ungroup()
}
```

## Figures for paper

### Figure 6: Compared to the name collection of Pubmed authors, honorees from North America are overrepresented while Europe & Central Asia and East Asia & Pacific honorees are underrepresented.

```{r fig.height=7, fig.width=9}
fig_5a <- iscb_pubmed %>%
  filter(year < '2020-01-01') %>%
  group_by(year, type, region) %>%
  summarise(mean_prob = mean(probabilities, na.rm = T)) %>%
  ungroup() %>%
  affiliation_breakdown('main', fct_rev(type)) +
  guides(fill = guide_legend(nrow = 2)) +
  theme(legend.position = 'bottom',
        legend.key.height = unit(3, 'mm'),
        legend.key.width = unit(3, 'mm'),
        legend.text = element_text(size = 8),
        legend.margin = margin(-0.2, 0.2, 0.2, 0, unit='cm'))

big_cat <- c('North America', 'Europe & Central Asia', 'East Asia & Pacific', 'Other Categories')
## Mean and standard deviation of predicted probabilities:
fig_5b <- iscb_pubmed_oth %>%
  filter((region %in% big_cat)) %>%
  loess_and_ci() +
  theme(legend.position = c(0.88, 0.83),
        legend.key.height = unit(4, 'mm'),
        legend.key.width = unit(5, 'mm'),
        legend.text = element_text(size = 6)) +
  facet_wrap(vars(fct_relevel(region, big_cat)), nrow = 1)

fig_5 <- cowplot::plot_grid(fig_5a, fig_5b, labels = 'AUTO', ncol = 1, rel_heights = c(1.3,1))
fig_5
ggsave('figs/geo_affiliation_breakdown.png', fig_5, width = 6.5, height = 5.5)

```

### Supplementary Figure S6 {#sup_fig_s6}

Proportion of authors from East Asia & Pacific had increased, particularly for papers published in Bioinformatics and BMC Bioinformatics.

```{r fig.width=6, fig.height=3.5}
fig_s6_1 <- bind_rows(pubmed_aff) %>%
  affiliation_breakdown(category = 'sub', journal) +
  theme(legend.position = 'right',
        legend.key.height = unit(3.5, 'mm'),
        legend.key.width = unit(2, 'mm'),
        legend.text = element_text(size = 7),
        legend.margin = margin(0, -0.03, 0, -0.2, unit='cm')
        )

  fig_s6_2 <- bind_rows(iscb_aff) %>%
  filter(year < '2020-01-01') %>%
  affiliation_breakdown(category = 'sub', journal)

fig_s6 <- cowplot::plot_grid(fig_s6_1, fig_s6_2, labels = 'AUTO', ncol = 1, rel_widths = 2)
fig_s6
ggsave('figs/fig_s6.png', fig_s6, width = 6, height = 3.5)
```

### Supplementary Figure S7 {#sup_fig_s7}

```{r fig.height=4, fig.width=7}
fig_s7 <- iscb_pubmed %>%
  loess_and_ci() +
  theme(legend.position = c(0.88, 0.25), legend.text = element_text(size = 7)) +
  facet_wrap(vars(fct_relevel(region, region_levels)), nrow = 2)
fig_s7
ggsave('figs/fig_s7.png', fig_s7, width = 7, height = 3.5)

```


```{r}
sessionInfo()
```

